---
title: "FZ_new"
output: html_document
date: "2026-08-07"
---

```{r }
library(tidyverse)    # Data cleaning and manipulation (includes dplyr, tidyr, etc.)
library(lubridate)    # Date and time handling
library(readxl)       # Read Excel files
library(glmmTMB)      # Fit Generalized Linear Mixed Models (GLMM)
library(broom.mixed)  # Tidy extraction of mixed-model parameters and CIs
library(DHARMa)       # Model residual diagnostics and hypothesis tests (zero-inflation, overdispersion)
library(MuMIn)        # Compute model R-squared
library(emmeans)      # Estimated marginal means
library(knitr)

# 1. Read camera deployment history table
camera_info <- read_excel("Fuzhou_Camera_info.xlsx")

# 2. Create a long table and compute "seasonal" effort (Season_Duration)
seasonal_effort <- camera_info %>%
  dplyr::select(Location, Camera, starts_with("Install"), starts_with("Check")) %>%
  pivot_longer(
    cols = starts_with(c("Install", "Check")),
    names_to = c(".value", "Period_ID"),
    names_pattern = "(Install|Check)(\\d+)",
    values_drop_na = TRUE
  ) %>%
  mutate(
    Install = ymd(Install),
    Check = ymd(Check)
  ) %>%
  filter(!is.na(Install) & !is.na(Check)) %>%
  # Expand each deployment period into a daily sequence
  mutate(Daily_Sequence = map2(Install, Check, ~seq(.x, .y, by = "day"))) %>%
  unnest(Daily_Sequence) %>%
  # Assign season labels
  mutate(
    Month = month(Daily_Sequence),
    Season = case_when(
      Month %in% c(3, 4, 10, 11, 12) ~ "Migration",
      Month %in% c(5, 6, 7, 8, 9) ~ "Summer",
      Month %in% c(1, 2) ~ "Winter"
    )
  ) %>%
  # Summarize seasonal effort
  group_by(Location, Camera, Season) %>%
  summarise(
    Season_Duration = n(), # Season effective monitoring days
    .groups = 'drop'
  )

camera_total_duration <- seasonal_effort %>%
  group_by(Camera) %>%
  summarise(Duration_1 = sum(Season_Duration), .groups = 'drop')

camera_info <- camera_info %>%
  left_join(camera_total_duration, by = "Camera")



```
# Calculate RAI (new)
```{r}
data <- read_excel("Data.xlsx")
speciescount <- read_excel("species_inventory.xlsx")
# Map species Chinese names to English names
data <- merge(data,speciescount,by.x="Species",by.y = "Chinese_Name",all.x = TRUE)
data_clean <- data %>% dplyr::select(Location, Camera, Common_Name, Taxonomy, Date, Number)

library(dplyr)
library(tidyr)

# 1. Remove invalid rows caused by missing taxonomy (Common_Name is NA)
data_clean <- data_clean %>%
  filter(!is.na(Common_Name))

data_clean <- data_clean %>%
  filter(Camera %in% camera_info$Camera)

# 2. Compute total monitoring duration for Fushan and Qishan
# Use Duration_1 from camera_info which was computed above
location_effort <- camera_info %>%
  group_by(Location) %>%
  summarise(Total_Duration = sum(Duration_1, na.rm = TRUE), .groups = "drop")

# 3. Weighted species counts by Location and Common_Name to derive new RAI
new_rai_calc <- data_clean %>%
  # Sum Number to get total recordings per species
  group_by(Location, Common_Name) %>%
  summarise(Weighted_Count = sum(Number, na.rm = TRUE), .groups = "drop") %>%
  # Join total monitoring time
  left_join(location_effort, by = "Location") %>%
  # Compute new RAI: (count / total_duration) * 100
  mutate(New_RAI = (Weighted_Count / Total_Duration) * 100)

# 4. Pivot to wide format to produce New_Fushan_RAI and New_Qishan_RAI
new_rai_wide <- new_rai_calc %>%
  dplyr::select(Common_Name, Location, New_RAI) %>%
  pivot_wider(
    names_from = Location,
    values_from = New_RAI,
    names_prefix = "New_",
    values_fill = list(New_RAI = 0)
  ) %>%
  rename(New_Fushan_RAI = New_Fushan, New_Qishan_RAI = New_Qishan)

# 5. Compare with old RAI in speciescount
rai_comparison <- speciescount %>%
  dplyr::select(Common_Name, Old_Fushan_RAI = Fushan_RAI, Old_Qishan_RAI = Qishan_RAI) %>%
  inner_join(new_rai_wide, by = "Common_Name") %>%
  dplyr::select(
    Common_Name,
    Old_Fushan_RAI, New_Fushan_RAI,
    Old_Qishan_RAI, New_Qishan_RAI
  ) %>%
  mutate(
    Old_Fushan_RAI = round(as.numeric(Old_Fushan_RAI), 2),
    New_Fushan_RAI = round(New_Fushan_RAI, 2),
    Old_Qishan_RAI = round(as.numeric(Old_Qishan_RAI), 2),
    New_Qishan_RAI = round(New_Qishan_RAI, 2)
  )

# View comparison
print(rai_comparison)

library(openxlsx)

# 1. Merge new RAI back into the original species table
speciescount_updated <- speciescount %>%
  # Defensive removal: drop if columns exist, harmless if they don't
  dplyr::select(-any_of(c("RAI_Fushan_n", "RAI_Qishan_n"))) %>%
  left_join(new_rai_wide, by = "Common_Name") %>%
  # 2. Rename to requested column names
  rename(
    RAI_Fushan_n = New_Fushan_RAI,
    RAI_Qishan_n = New_Qishan_RAI
  ) %>%
  # 3. Replace NA with 0 for species not recorded in this run
  mutate(
    RAI_Fushan_n = replace_na(RAI_Fushan_n, 0),
    RAI_Qishan_n = replace_na(RAI_Qishan_n, 0)
  )

# Inspect
head(speciescount_updated)

# 4. Optionally overwrite the original Excel file
#write.xlsx(speciescount_updated, "species_inventory.xlsx", rowNames = FALSE)

cat("New RAI columns have been written and exported to species_inventory.xlsx.\n")

```
# Calculate distances between cameras
```{r}
# Load required packages
library(tidyverse)
library(readxl)
library(geosphere) # For spherical distances between lat/lon
library(knitr)

# 1. Read and clean camera info
camera_info <- read_excel("Fuzhou_Camera_info.xlsx")

cam_coords <- camera_info %>%
  # Remove non-numeric characters from coordinates (e.g., °N, °E) and convert to numeric
  mutate(
    Latitude = as.numeric(str_replace_all(Latitude, "[^0-9.]", "")),
    Longitude = as.numeric(str_replace_all(Longitude, "[^0-9.]", "")),
    Elevation = as.numeric(Elevation)
  ) %>%
  dplyr::select(Location, Camera, Longitude, Latitude, Elevation) %>%
  drop_na(Latitude, Longitude, Elevation)

# 2. Define function to compute nearest-neighbor distances (NND)
get_nn_distances <- function(df) {
  # If fewer than 2 cameras in site, return empty numeric
  if(nrow(df) < 2) return(numeric(0)) 

  # A. 2D spherical distance matrix (meters)
  dist2d_mat <- distm(df[, c("Longitude", "Latitude")])

  # B. Elevation difference matrix (meters)
  elev_mat <- as.matrix(dist(df$Elevation))

  # C. Combine via Pythagorean theorem to approximate 3D distance
  dist3d_mat <- sqrt(dist2d_mat^2 + elev_mat^2)

  # D. Avoid self-distance being selected as min
  diag(dist3d_mat) <- NA

  # E. Find minimum for each row
  nn_dists <- apply(dist3d_mat, 1, min, na.rm = TRUE)

  return(nn_dists)
}

# 3. Compute nearest-neighbor distances per site
fushan_nn <- get_nn_distances(cam_coords %>% filter(Location == "Fushan"))
qishan_nn <- get_nn_distances(cam_coords %>% filter(Location == "Qishan"))

# Combine
overall_nn <- c(fushan_nn, qishan_nn)

# 4. Summary statistics: mean and SD
distance_summary <- tibble(
  Scope = c("Fushan", "Qishan", "Overall"),
  Mean_Nearest_Distance_m = c(mean(fushan_nn), mean(qishan_nn), mean(overall_nn)),
  SD_Nearest_Distance_m = c(sd(fushan_nn), sd(qishan_nn), sd(overall_nn)),
  Camera_Count = c(length(fushan_nn), length(qishan_nn), length(overall_nn))
) %>%
  mutate(
    Mean_Nearest_Distance_m = round(Mean_Nearest_Distance_m, 2),
    SD_Nearest_Distance_m = round(SD_Nearest_Distance_m, 2)
  )

# 5. Print results
kable(distance_summary, format = "markdown")
```

# GLMM modeling
```{r}
library(glmmTMB)
library(MuMIn)
data <- read_excel("Data.xlsx")
speciescount <- read_excel("species_inventory.xlsx")
# Map species Chinese names to English names
data <- merge(data,speciescount,by.x="Species",by.y = "Chinese_Name",all.x = TRUE)
data <- data %>% filter(Taxonomy %in% c("Bird","Mammal"))
data_clean <- data %>% dplyr::select(Location, Camera, Common_Name, Taxonomy, Date, Number)

data_clean <- data_clean %>%
  filter(!Camera %in% c("FT10", "FT11", "FT12", "FT13", "FT14")) %>%
  mutate(Camera = case_when(
    Camera == "QT7" ~ "Q7",
    Camera == "QT10" ~ "Q10",
    TRUE ~ Camera
  )) %>%
  arrange(Camera)

glmm_input_seasonal <- data_clean %>%
  filter(!Common_Name %in% c("bat", "rat", "shrew", "leaf warbler")) %>%
  # Assign season (must match season calculation exactly)
  mutate(
    Date = as.Date(Date),
    Month = month(Date),
    Season = case_when(
      Month %in% c(3, 4, 10, 11, 12) ~ "Migration",
      Month %in% c(5, 6, 7, 8, 9) ~ "Summer",
      Month %in% c(1, 2) ~ "Winter"
    )
  ) %>%
  # Count species per season
  group_by(Location, Camera, Taxonomy, Season) %>% 
  summarise(
    Species_Number = n_distinct(Common_Name[Common_Name != "No_Species_Detected"]),
    .groups = 'drop'
  ) %>%
  # Join seasonal effort (right_join to keep all actual working seasons)
  right_join(seasonal_effort, by = c("Location", "Camera", "Season")) %>%
  # Replace NA Species_Number (no detections) with 0
  mutate(Species_Number = replace_na(Species_Number, 0)) %>%
  # Ensure each worked Camera+Season has both Mammal and Bird rows
  complete(nesting(Location, Camera, Season, Season_Duration), Taxonomy, fill = list(Species_Number = 0)) %>%
  # Remove any accidental NA Taxonomy rows
  filter(!is.na(Taxonomy)) %>%
  # Keep only records with Season_Duration > 0 to avoid log(0)
  filter(Season_Duration > 0)
```

# Model fitting
```{r}
library(glmmTMB)

# Fit GLMM
model_season_glmm <- glmmTMB(
  Species_Number ~ Location + Taxonomy + Season + offset(log(Season_Duration)) + (1|Camera),
  data = glmm_input_seasonal, 
  family = nbinom2 # Negative binomial for overdispersion
)

summary(model_season_glmm)
```

model_season_glmm <- glmmTMB(
  Species_Number ~ Location + Taxonomy + Season + offset(log(Season_Duration)) + (1|Camera),
  data = glmm_input_seasonal, 
  family = poisson(link = "log")
)

model_season_glmm <- glmmTMB(
  Species_Number ~ Location + Taxonomy + Season + log(Season_Duration) + (1|Camera),
  data = glmm_input_seasonal, 
  family = poisson(link = "log")
)

model_season_glmm <- glmmTMB(
  Species_Number ~ Location + Taxonomy + Season + log(Season_Duration) + (1|Camera),
  data = glmm_input_seasonal, 
  family = nbinom2
)

model_season_glmm <- glmmTMB(
  Species_Number ~ Location* Taxonomy +Season+ log(Season_Duration) + (1|Camera), 
  data = glmm_input_seasonal, 
  family = poisson(link = "log")
)

#the chosen one
model_season_glmm <- glmmTMB(
  Species_Number ~ Location* Taxonomy +Season* Taxonomy+ log(Season_Duration) + (1|Camera), 
  data = glmm_input_seasonal, 
  family = poisson(link = "log")
)

summary(model_season_glmm)
```


```{r}
tidy_results <- tidy(model_season_glmm, conf.int = TRUE, effects = "ran_pars")
print(tidy_results)
(exp(0.17436422)^2)

model_ci <- confint(model_season_glmm, method = "profile")
print(model_ci)

# model's % variantion explained (R-squared) 
r2_overall <- r.squaredGLMM(model_season_glmm)
print(r2_overall)

simulated_res <- simulateResiduals(fittedModel = model_season_glmm, plot = TRUE)

testDispersion(simulated_res)       
testZeroInflation(simulated_res)     

# ==============================================================================
# 5. emmeans to report model-predicted values
# ==============================================================================
pred_emmeans <- emmeans(model_season_glmm, 
                        specs = ~ Location + Taxonomy, 
                        at = list(Season_Duration = 100), 
                        type = "response")

print("===== model-predicted number of species in 100 days =====")
print(pred_emmeans)

library(emmeans)

# 1. each combinitaion's marginal means (Season_Duration = 100）
pred_emmeans_int <- emmeans(model_season_glmm, 
                            specs = ~ Taxonomy | Location, 
                            at = list(Season_Duration = 100), 
                            type = "response")

# 2. print predicted values
print(pred_emmeans_int)

# 3, Post hoc comparison
pairwise_comparisons <- pairs(pred_emmeans_int)
print(pairwise_comparisons)

```
